\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-05-27 paper/ms.tex
adding full specifications for the models

After computing the Cholesky factorization of the matrix $K$, we can apply
the inverse of $K$ and compute its log-determinant in $\mathcal{O}(N\,J)$ and
$\mathcal{O}(N)$ operations respectively.
The log-determinant of $K$ is given by
\begin{eqnarray}
    \log \det K = \sum_{n=1}^N \log D_{n,n}
\end{eqnarray}
where, if $K$ is positive semi-definite, $D_{n,n}>0$ for all $n$.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/sho.pdf}
\caption{(left) The power spectrum of a stochastically-driven simple harmonic
    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
    factor $Q$.
    For comparison, the dashed line shows the Lorentzian function from
    \eq{lorentz-psd} with $c_j = \omega_0/(2\,Q) = 1/20$ and normalized so that
    $S(d_j)/S(0) = 100$.
    (right) The corresponding autocorrelation functions with the same colors.
    \figurelabel{sho}}
\end{center}
\end{figure}

@@

@@

@@

A source of variability that can be measured from time series measurements of
stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the rotation period of the star.
GP modeling with the same kernel function has been proposed as a method of
measuring the mean periodicity in quasiperiodic photometric time series in
general \citep{Wang:2012}.
The key difference between \eq{sine2} and other quasiperiodic kernels is that
it is positive for all values of $\tau$.
We can construct a simple \celeriteterm\ covariance function with similar properties
as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{a}{2+b}\,e^{-c\,\tau}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
\right]
\end{eqnarray}
for $a>0$, $b>0$, and $c>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a substantial gain in computational efficiency.

@@

@@

@@

@@

@@

\acknowledgments
It is a pleasure to thank
Megan Bedell,
Ian Czekala,
Will Farr,
Sam Grunblatt,
David W.\ Hogg,
Dan Huber,
Meredith Rawls,
Dennis Stello,
Jake VanderPlas, and
Andrew Gordon Wilson
for helpful discussions informing the ideas and code presented here.

>>>>

After computing the Cholesky factorization of the matrix $K$, we can apply
the inverse of $K$ and compute its log-determinant in $\mathcal{O}(N\,J)$ and
$\mathcal{O}(N)$ operations respectively.
The log-determinant of $K$ is given by
\begin{eqnarray}
    \ln \det K = \sum_{n=1}^N \ln D_{n,n}
\end{eqnarray}
where, if $K$ is positive semi-definite, $D_{n,n}>0$ for all $n$.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.95\textwidth]{figures/sho.pdf}
\caption{(left) The power spectrum of a stochastically-driven simple harmonic
    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
    factor $Q$.
    For comparison, the dashed line shows the Lorentzian function from
    \eq{lorentz-psd} with $c_j = \omega_0/(2\,Q) = 1/20$ and normalized so that
    $S(d_j)/S(0) = 100$.
    (middle) The corresponding autocorrelation functions with the same colors.
    \response{(right) Three realizations from each model in the same colors.}
    \figurelabel{sho}}
\end{center}
\end{figure}

@@

\input{figures/simulated/correct-params}

@@

\input{figures/simulated/wrong-qpo-params}

@@

A source of variability that can be measured from time series measurements of
stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the rotation period of the star.
GP modeling with the same kernel function has been proposed as a method of
measuring the mean periodicity in quasiperiodic photometric time series in
general \citep{Wang:2012}.
The key difference between \eq{sine2} and other quasiperiodic kernels is that
it is positive for all values of $\tau$.
We can construct a simple \celeriteterm\ covariance function with similar properties
as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{a}{2+b}\,e^{-\tau/c}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
\right]
\end{eqnarray}
for $a>0$, $b>0$, and $c>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a substantial gain in computational efficiency.

@@

\input{figures/rotation/rotation-params}

@@

\input{figures/astero/astero-params}

@@

\newpage
\input{figures/transit/transit-params}

@@

\subsection{Summary}

\input{figures/example-stats}

@@

\acknowledgments
It is a pleasure to thank
Megan Bedell,
Ian Czekala,
Will Farr,
Sam Grunblatt,
David W.\ Hogg,
Dan Huber,
Meredith Rawls,
Dennis Stello,
Jake VanderPlas, and
Andrew Gordon Wilson
for helpful discussions informing the ideas and code presented here.
\response{We would also like to thank the anonymous referee for thorough and
constructive feedback that greatly improved the paper.}
